TacticToe: Learning the Game of Tactics

### 2018-01-27 tactictoe_jv.tex
removing comments

%
%\begin{abstract}
%Formalization of mathematical statements in interactive theorem provers (ITPs) 
%relies nowadays on many methods to automate the construction of a proof.
%Some of this method are general purpose and with enough time can prove any 
%theorems. Yet, in a formalized proof script 
%a mathematician take advantages of a combination of specialized rules, 
%theory-based strategy and general ones. All of them are implemented as tactics 
%in HOL4. Automated theorem provers (ATPs) rely on parameterizable general 
%=======
\todo{JU review: Still feels more like an intro than an abstract. Could be 
shorter and punchier.}
\begin{abstract}
Formalization of mathematics in interactive theorem provers (ITPs) 
relies today on many proof automation methods.
Some of these methods are general-purpose and with enough time they can prove any 
theorem. Many formalized proof scripts however also
take advantage of specialized rules and theory-based strategies. 
In the HOL4 ITP system, practically all of them are implemented as \emph{tactics}.

@@

We bridge this gap by implementing our tactical prover TacticToe on top 
of the HOL4 ITP. 
TacticToe first learns from many proof scripts
the relevance of tactics in particular proof situations.
% are effective from 
% tactic-level proofs and what kind of formulas are the results of effective 
% steps, 
This learned knowledge is then used in a Monte Carlo tree search algorithm that explores the most promising tactic-level proof paths. 
TacticToe proves 66.2\% of 7164 theorems of the standard library where as 
Eprover solves 34.5\%. The success rate rises to 69.0\% by combining the 
result of both provers.

%on their first-order translations?

%
%Techniques combining machine learning with translation to automated
%reasoning have recently become an important component of formal proof 
%assistants. 
%Such ``hammer'' techniques complement traditional proof
%assistant automation as implemented by tactics and decision
%procedures. n this paper we present a unified proof assistant
%automation approach which attempts to automate the selection of
%appropriate tactics and tactic-sequences combined with an optimized
%small-scale hammering approach.
%
%We implement the technique as a tactic-level automation for HOL4:
%TacticToe. It implements a modified A*-algorithm directly in HOL4 that
%explores different tactic-level proof paths, guiding their selection
%by learning from a large number of previous tactic-level
%proofs. Unlike the existing hammer methods, TacticToe avoids
%translation to FOL, working directly on the HOL level. By combining tactic 
%prediction and premise selection, TacticToe is able to re-prove 39\% of 7902 
%HOL4 theorems in 
%%less than
%5 seconds whereas the best single HOL(y)Hammer strategy solves 32\% in the 
%same 
%amount of time.

\end{abstract}

\section{Introduction}\todo{TG + JU + CK}
%\begin{example} Proof automatically generated by \tactictoe for the user given 
%goal
%\small
%\begin{center}
%
%
%\begin{BVerbatim}[commandchars=\\\{\}]
%Goal: ``\(\forall\)l. FOLDL (\(\lambda\)xs x. SNOC x xs) [] l = l``
%Proof:
%  SNOC_INDUCT_TAC THENL
%  [ REWRITE_TAC [APPEND_NIL, FOLDL],
%    ASM_REWRITE_TAC [APPEND_SNOC, FOLDL_SNOC]
%      THEN CONV_TAC (DEPTH_CONV BETA_CONV)
%      THEN ASM_REWRITE_TAC [APPEND_SNOC, FOLDL_SNOC] ]
%\end{BVerbatim}
%
%
%\end{center}
%
%\end{example}
%

\tableofcontents

\todo{TG : comparison with hammers and advantages for the user and over hammers}
Many of the state-of-the-art interactive theorem provers (ITPs) such as
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight}, 
  \isabelle~\cite{isabelle} 
  and \coq~\cite{coq-book}
  provide high-level parameterizable tactics for constructing the
  proofs.  Such tactics typically analyze the current goal state and
  assumptions, apply nontrivial proof transformations, which get
  expanded into possibly many basic kernel-level inferences or significant 
  parts of the proof term.
%\marginpar{not Coq}
  In this
  work we develop a tactic-level automation procedure for the \holfour ITP
  which guides selection of the tactics by learning from previous
  proofs.  Instead of relying on translation to first-order automated
  theorem provers (ATPs) as done by the hammer 
  systems~\cite{hammers4qed,tgck-cpp15}, the technique
%\marginpar{too many cites?}
%BlanchetteGKKU16,holyhammer,mizAR40
  directly searches for sequences of tactic applications that lead to
  the ITP proof, thus avoiding the translation and
  proof-reconstruction phases needed by the hammers.

\todo{TG: rewrite the plan}
  To do this, we \emph{extract and record} tactic invocations from the ITP 
  proofs (Section~\ref{sec:recording}) and
  \emph{build efficient machine learning classifiers} based on such training
  examples (Section~\ref{s:prediction}).  The learned data serves as a 
  guidance for a \emph{Monte Carlo tree search} that explores the different 
  proof paths (Section~\ref{sec:proofsearch}). The result, if
  successful, is a certified human-level proof composed of \holfour
  tactics.  The system is evaluated on a large set of theorems originating
  from \holfour (Section~\ref{s:experiments}), and we show that the performance 
  of the single best \tactictoe strategy exceeds the performance of a hammer 
  system used with a single strategy and a single efficient external prover.
%Thanks to a very different proof-search procedure, \tactictoe also
%proves many theorems whose proofs are not found by the hammering
%approach.

\subsection{The problem}\todo{ }

How a user works with an ITP and how this gives rise to a machine learning problem.

@@

@@

\todoi{add prediction of 4 kinds of things}
\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern 
matching constructions, opening of submodules are supported. The different 
recording steps (parsing, unfolding and replaying) are regrouped into a single 
\sml function. So, our recording is more complete and fully automatized.
%\item Existing tactic orthogonalization on recorded data is made stronger by 
%also including the internal ATP \metis in the orthogonalization 
%competition. Because we manage a database of theorems for argument 
%prediction, we apply here a pruning step to remove easy theorems.
\item Two types of arguments (list of theorems and terms) can be predicted 
independently of the tactic. Combinatorial explosion in the number of tactics 
is prevented through the introduction of abstracted tactics. As desired, their 
instantiations is dependent on the target goal.
\item MCTS (Monte Carlo tree search) now replaces A* in our proof search 
algorithm. MCTS gives a more balanced feedback to guide proof search by 
comparing subtrees instead of leafs. The policy and evaluation are learned 
through supervised learning.

@@

\section{Prediction}\label{s:prediction}\todo{TG: complete term theorem and 
list of goals specificities in terms of predictions and justify by their use 
case}

The learning-based selection of relevant lemmas significantly improves the 
automation for hammers~\cite{BlanchetteGKKU16}. In \tactictoe, we adapt 
one of the strongest lemma selection 
methods: the modified distance-weighted \emph{$k$ nearest-neighbour} 
classifier~\cite{ckju-pxtp13,DudaniS76} to predict four different objects: 
terms, goals, theorems and list of goals.

Before going into the details of the machine learning algorithm, we introduce
specificities associated with the prediction of each object: the data from 
which the feature vectors are build and the different use cases for the 
prediction.

We derive from tactic selection a prior policy for 
our search algorithm. Lists of goals (products of tactics) evaluate the success 
of the policy and adapt it accordingly. Term and theorems predictions are 
necessary to find the suitable arguments for abstracted tactics (see Section 
\ref{sec:synthesis}).

%Premise selection usually only prunes the initial set of formulas given to the 
%ATPs, which then try to solve the pruned problems on their own. 
%the goal of predicting useful lemmas is to reduce the search space,
% of the external ATPs called by a hammer,

%Mention explictly all the predictors that will be using this distance.
%tactics:
%theorems:
%terms:
%goal list: (product of tactics)
\paragraph{term}
Predict the closest term in a tactic that appears in the goal.

\paragraph{theorem}
Predict the theorem the closest to the current goal. Theorem dependencies are 
added during preselection.

\paragraph{tactic via goal}
Given a goal $g$, the classifier selects a list of previously solved goals $l$
ordered by their similarity to $g$. Since each solved goal is associated 
with a tactic in our database, the ordered list of goals induces a ordered  
list of associated tactics. And tactics are tried in this order during the 
proof search. Therefore, to order the tactics correctly, it is crucial to 
estimate the similarity between goals with mathematically relevant features.

@@

Whenever we mention that we want to predict a goal, we 
actually intend to predict its associated tactics.

\paragraph{list of goals}
Used for Monte Carlo Tree Search evaluation. Predict the closest list of goals
to the list of goal produced by a tactic.

%We will next discuss 
%the extraction 
%of features from goals and the
%actual prediction of tactics. 
%Both have been integrated in the \sml proof search.

\subsection{Feature Extraction}\label{sec:features}
The four objects considered are different representation of mathematical 
formulas. Therefore, we start by describing which features we extract from the 
simplest representation terms. From there, we extend the extraction mechanism 
to goals theorems and lists of goals. Duplicated features are always removed 
so each object has an associated set of features.

@@

%We found that the names of variables present in a term are particularly 
%important for predicting tactics such as 
%case splitting on a variable 
%(\texttt{Cases\_on var}) or instantiation of a variable 
%(\texttt{SPEC\_TAC var term}). 
%We additionally extract the following features:

%\begin{itemize}
%\item the top-level logical structure with atoms substituted by a single place 
%holder $A$ and all its substructures.
%\end{itemize}

%Determining the presence of top-level logical 
%operators (i.e 
%implication) is essential to assess if a "logical" tactic should be applied. 
%For example, 
%the presence of an implication may lead to the application of the tactic 
%\texttt{DISCH\_TAC} that moves the precondition to the assumptions. Top-level 
%logical structure gives a more detailed view of the relationship between those 
%logical components.

@@

Making prediction for an object $o$ consists sorting a set of objects 
by their similarity to $o$.
\todo{mention preselection and which dataset of objects?}

We use $sim_1$ for goals and theorems and $sim_2$ for terms and list of goals.
The similarity $sim_1$ is used for predicting a tactic (via goals) because if a
tactic is effective on large goal $g$ then it is often suitable for a goal $g'$ 
containing sub-formulas of $g$. The same monotonicity heuristic can justify the 
use of $sim_1$ in theorems. Indeed, in practice, a large theorem is often a 
conjunction of small theorems from the same domain. And if a conjunction 
contains a formula related to a problem, then the other conjuncts 
from the same domain may also help.

%\[tactic\_score_2 (f_o,f_p) = \frac{tactic\_score_1 (f_o,f_p)}{(1+ ln\ (1 + 
%card\ f_o))} \]

%Moreover, we would like to compare the distance of a recorded goal to 
%different goals opened at different moment of the search. That is why we 
%normalize the scores by dividing them by the similarity of the open goal with 
%itself. As a result, every score will lie in the interval $[0, 1]$ where 1 is 
%the highest degree of similarity (i.e. the shortest distance). We respectively 
%refer to those normalized scores later as $tactic\_norm\_score_1$ and 
%$tactic\_norm\_score_2$.

\subsection{Preselecting feature vectors and their  
recursive dependencies}\label{sec:dependencies}
We first rank feature vectors by their distance to the conjecture. But better 
ranked vectors only contains symbols from 
the conjecture and may not be suitable for later stages of the proof where the 
goals may have derived significantly from the conjecture. To solve this 
problem, we reorder the list so that descendants (see 
Definition~\ref{def:recdep}) of vectors inherits the similarity of their 
parent.

@@

%In practice, such preselection is necessary because our nearest neighbor 
%algorithm  is consuming more than half of the proof search time when there were
%more than 50,000 feature vectors to select from.

%\begin{remark}
%A preselection technique is also in place for the 
%theorem predictor and the evaluation predictor defined later in this paper, 
%because those predictors are also called during the proof search.
%\end{remark}

@@

\subsection{Argument Predictors}\label{sec:arg_pred}
We made argument predictors for two types of arguments: list 
of theorems and term. We choose this two types because they are the most common 
inside tactics and because perfect synthesis of those arguments would lead to 
the largest improvements (see Section~\ref{sec:perfect_exp}). Arguments with 
different types 
are currently never abstracted in tactics since we do 
not have yet a way to predict those arguments and instantiate those tactics.

\paragraph{List of Theorems}
Predicting theorems is made through the same system as premise selection
used for \holyhammer for general purpose automation which produces a list of 
theorems $l$. The original list of theorem $l'$ abstracted in the tactic is 
forgotten. For the abstracted tactic 
to replace the original one, the assumption is that the tactic does not break 
when adding arguments. If this condition is met, the abstracted tactic will 
always subsumes the original tactic when $l$ is a superset of $l'$. Such 
condition is called monotonicity of the 
tactic in its argument. Predicting such superset may not be necessary as the  
predicted theorems can sometimes be even more suitable than the original one 
because they depend directly on the current goal. One of the biggest advantage 
of predicting theorems independently is that tactics may now rely on theorems 
that were never part of a tactic.

\paragraph{Term}
Predicting terms is harder since we do not have a list of terms to choose from.
As an approximation for this list, we take the list of subterms $l$ contained 
in a goal $g$. Even then choosing the right subterms require us to look at the 
term $t$ the tactic argument was abstracted from. The term predictor then 
selects the subterm in $l$ that is closest to $t$. In an attempt to predict 
variables (which are terms in \holfour) at the place where they were previously 
variable, we use the distance $sim_2$ that penalizes large subterms.
Since we rely on the original term argument $t$, the kind of tactics produced 
are much less general that for abstracted list of terms and thus have smaller
coverage.

%In the orthogonalization settings, the abstracted forms compete alongside the 
%originals and are ranked according to their frequency. 
%Freq (abstract) = max (orig,abstr). And the abstracted from is always 
%preferred 
%to the original form of a tactic if both had the effect that was fought upon.
%
%
%Abstracted tactics then become part of the feature space and can be predicted 
%by the tactic predictors but are slightly slower because of the added overhead
%of predicting list of theorems for their unspecified arguments.
%
%We first detect if the tactics contains one or more theorem list. This 
%detection
%is done by looking if the first element of the list is of type thm.
%
%This method is difficult to generalize to any tokens?
%
%By forgetting what was the original argument, we lose an important piece of 
%information on what kind of arguments a tactic accepts. This is less problem 
%for
%a list of theorems because predicted theorems that are not usable by the 
%tactic 
%are usually ignored.
%
%What if you would have to predict a single theorem, or term, or function?
%
%There are two complementary guiding methods to predict them:
%
%- Dependent on the goal.
%- Similar to previous arguments for this tactic.
%
%For example one way to predict subterms, would be to take the subterm that is 
%the closest to predicted subterm that appears on the goal.
%
%For that we take slight variation of the previously defined distance to 
%eliminate large subterms:

%\subsubsection{Estimating the difficulty of a goal}
%A first idea is to guess what is the expected length of its shortest proof.
%
%From the recorded feature vectors, we can compute the length of the shortest 
%proof for each goal $g$ and can be defined recursively by:
%
%length :::= mathematical stuff
%
%Now to compute the difficulty of a goal. We can average the difficulty of the 
%$k$ nearest neighbor. 
%
%However, we believe that it is more important to try to guess if a goal is 
%likely to be provable or not.
%
%We augment our data by looking at all the produced list of goals
%that did not subsumes the original one during 
%orthogonalization. We had the list to our list of negative examples. The list 
%of positive examples are the goal list that are present in the feature vectors.
%
%
%We try to estimate if a goal was useful by taking the k nearest neighbor of 
%the 
%goal and see how many appeared in a previous proofs and how many are hard 
%negatives. The average give us a probability on how provable it is.
%We multiply this probability by a coefficient k.
%
%Using this heuristic to guide our proof search, we hope to reduce the amount 
%of 
%time that the search spends trying to prove unprovable goals.
%
%Formula for the heuristic.

%\begin{remark}
%In practice, each proof node contains a list of goals to be proven and we 
%estimate the difficulty of proving a conjunction of goals.
%\end{remark}

@@

\subsection{A search tree: data structures and definitions}

@@

Maybe should move this defnition in static description.

@@

The second case happens very rarely (less than 10 times when evaluating 7088 
theorems) and only at the beginning of the library when the learned dataset is 
small. So in our experiments, we just give the number of proof found relative 
to the number of proof attempts.

\end{definition}

%A large difference between diagonalized search and our MCTS algorithm is in 
%what is considered to be the width of the tree. In diagonalized search, 
%non-productive tactics contribute to the extension of the width where as in 
%MCTSn the width includes only the total number of children produced. The 
%influence of non-productive on the proof exploration is recovered when we 
%include an 
%evaluation function that takes into account the productivity of tactics such 
%as 
%$E_{productive}$ or $E_r$.

%Obviously, we can consider the 
%trivial evaluation function where 
%a goal is given reward if and only if it is proven. Yet, to make this approach 
%effective would require to learn on a very large amount of relatively easy 
%problems. To lessen the requirement of an very large amount of data, we 
%propose 
%to use our goal distance to select the $k$ closest goals that were proven 
%before. The following formulas shows how we can combine positive examples and 
%negatives examples to evaluate an unseen 
%goals.
%
%%Hard Negative mining (cite Deep math)
%
%\[\mathfrak{N}_m=\lbrace g\ |\ (N(g) \wedge n(g) \geq m) \vee 
%(P(g) \wedge p(g) >  m)\rbrace\]
%\[\mathfrak{P}_m=\lbrace g\ |\ P(g) \wedge p(g) \leq m \rbrace\]
%
%
%\[E(m) = \frac{1}{m} \times \frac{card\ \mathfrak{P}_m}
%    {card\ \mathfrak{N}_m + card\ \mathfrak{P}_m} \]
%
%
%\[E = \argmax_{m \in \mathbb{N}} E(m)\]
%

%We need
%MCTS.
%We use supervised learning to estimate evaluation functions.
%
%Positives are given the number of intermediate steps needed to finish the 
%proof.
%And the distance for negatives is two times the number of intermediate steps 
%tried. If we only tried one more tactic on this goal we would solve it.

%We use K-nn on them with k = 10 and calculate an the following weighted mean.
%
%
%The heuristic component should represent the number of steps (or total time) 
%estimated required to 
%solve the goal.
%If a goal is not provable then it's the number of time the goal was visited 
%during the search + 1.

%
%
%Combining with general automation:
%metis.
%holyhammer. (with orthogonalized theorems).
%divide by the sum of the weight of all features.
%
%
%New astar heuristic:
%The astar algorithm was simplified and now is represented in the standard 
%additive form instead of the multiplicative one described in our previous 
%paper.
%
%We also tried to estimate the difficulty of a goal. How likely that a 
%conjunction of goals are provable.
%
%Heuristic from prediction score:
%
%Heuristic from positives:
%
%Heuristic from positives + hard negatives.
%
%Admissibility of the heuristic:
%  
%Connection with UCB, UCT and Monte Carlo.
%
%Why is it not applicable in our case?
%
%1) Monte carlo requires to explore every action at least once. 
%This could be fixed by predicting fewer tactics 10-20. But we do not have
%yet an automatic way to tell.
%
%2) Monte Carlo requires a evaluation of the provability of the goal.
%This is insight can be gain.
%
%\subsection{Node selection formula}
%Monte Carlo tree search is a tree search algorithm. It explores the proof tree
%by performing simulation from the root to a leaf of the tree by selecting each 
%node by looking at that includes exploration and exploitation. Intuitively, 
%the 
%trade-off between exploitation and exploration means that the algorithm will 
%exploit deeper more promises branches but leaving enough time for exoloring 
%less likely alternatives.
%
%
%
%Connection with previous tree search. The prior probabilities are decreasing
%exponential of 2s. $\sqrt (n) > 1/2^n$
%
%
%
%
%
%calls simulation. 
%
%
%
%\subsection{Description of the evaluation function}
%
%Non local heuristic for evaluation. Should consider also what is needed to be 
%proven.
%
%
%In order to evaluate how long it takes to prove a goal. We collect data from 
%previous proof searches. Those include negative and positive examples.
%Positives examples are associated with the minimal number of steps needed to 
%find their proofs. Negative examples carry along the largest proof search 
%unsuccessfully performed.
%
%Combining this measure in an single evaluation of the proof length is a 
%challenging task. We use the following formulas P1 for estimating the 
%probability of being provable in less than one step.
%
%Neg(i) = not provable in i steps.
%70 percent change of not being provable in two steps.
%50 percent chance of closing the proof in 1 step: 2
%1 percent chance of being provable in 1 steps.
%20 percent chance of being provable in 2 steps. 5*2 = value = 10
%100 percent chance of being provable in 5 steps. 5
%
%\[P_i = Neg(i) | Knn(10) \] 
%
%the total estimation is given by:
%\[E = (\sum_{i=1}^5 (1-P_i)) + 1\] 
%
%\subsection{Combining evaluation function on goals}
%
%Independence of each probabilities.
%Full correlation: $(P1 * P2 * P3) ^{1/3}$
%Min probabiity:

%
%\subsection{Advantage of Monte Carlo Tree Search over A*-search}
%- Take into account all estimation along the proof and not only the last leaf.
%- Easy to link conjunction estimation   together.

\subsection{ATP Integration}
General-purpose proof automation mechanisms which combine proof translation to
ATPs with machine learning (``hammers'') have become quite successful in
enhancing the automation level in proof assistants~\cite{hammers4qed}.
As external automated reasoning techniques sometimes outperform the combined 
power of tactics, we would like to combine the \tactictoe search with 
stronger general purpose provers such as the one in \holyhammer for 
\holfour~\cite{tgck-cpp15}.

During our experiments,
We chose to distinguish two these general purpose tactic \metis and \eprover.
\metis is integrated in \holfour, so can be call with frequently with low 
timeout.
And the strongest external provers according to past experiments \cite{tgck-cpp15}.
Where as...
external calls to \eprover are computationally expensive (5 seconds at a 
minimum +  128 premises), we  
experiment with a small hammer tactic comprised of: a faster premise 
selection algorithm and
a short call (0.1 seconds + 16 premises optimized by the previous paper) to the 
internal prover \metis~\cite{metis} or a
longer call to external prover \eprover.

@@

%The number of parallel threads can be capped.
%And if tactictoe or an asynchronous thread proves a goal, all descendants of 
%these goal are deactivated, terminating associated threads. This frees space 
%for more pending threads, associated to newly created nodes, to start.

%needs to be taken after each proof step to close proofs solved by \eprover
%When the number of thread reaches a limit, a ``hammer'' tactic call 
%should
%wait for a thread to be freed but the proof search goes on and the feedback of 
%the ``hammer'' tactic is delayed. 
%
%
%It is best tactic available but we do not have any data about it since it does 
%not appear in any proofs. So , we 
%artificially try it on

%\paragraph{Effect on orthogonalization of tactics}
%During orthogonalization, the general purpose tactic will be the first 
%competitor and if it manages to prove the goal it will subsume the original 
%tactic on this goal since the empty set of goals is a subset of any set. 
%Therefore, it will replace the original tactic in the feature vector that is 
%competed for. This is desirable since we would like to remove alternative 
%proofs.

%\paragraph{Orthogonalization of theorems}
%A ``hammer'' tactic can be useful to remove parallel
%theorems. We consider a theorem to be parallel if it can be proven 
%from previous theorems using a general tactic with a timeout equal or less 
%than the one it is given during the proof search. Parallel theorems are marked 
%and removed from the set of theorems $\mathfrak{T}$ from 
%which premise selection is performed.
%
%\begin{remark}
%The ``smaller hammer'' is chosen to remove parallel theorems and as the first 
%try during orthogonalization of tactics. 
%\end{remark}

%\paragraph{Effect on the proof search}
%First, before the proof search, we preselect 500 theorems 
%for the whole proof search tree using a premise selection algorithm 
%that takes into account dependencies between theorems.
%During the proof search, when a new goal is created or a fresh pending goal is 
%considered, the ``hammer'' tactic is always tried first because if it succeeds 
%the node would be closed rapidly.

%\paragraph{Parallelization of external calls}
%It is possible to parallelize calls to external provers so that they happen 
%asynchronously during the proof search. We let the prover runs \tactictoe 
%continue its proof in parallel of a ``hammer'' tactic call.

@@

\begin{table}[ht]
\centering\ra{1.3}
\small
\begin{tabular}{ll}
\toprule
 $c_{policy}$ & Solved \\
\midrule
 $0.66$ & 120\\
 $0.5$ & 125\\
 $0.33$ & 117\\
% 1/2 & Productive & No & $E_{none}$ & 1 & 120\\
% 1/2 & Productive & No & $E_{productive}$ & 1 & 127\\
\bottomrule
\end{tabular}
\caption{\label{tab:cost_param} 
%on a local machine
}
\end{table}

@@

This experiments are performed with orthogonalization on and policy coefficient 
of $0.5$.

\begin{table}[ht]
\centering\ra{1.3}
\small
\begin{tabular}{lll}
\toprule
 Evaluation & $c_{exploration}$ & Solved \\
\midrule
% 1/2 & All & No & $E_{none}$ & 1 & 122\\
% 1/2 & All & Yes & $E_{none}$ & 1 & 128\\
 $E_{none}$ & 1 & 128\\
 $E_{productive}$ & 1 & 134\\ 
 & 4 & 130\\
 $E_1$ & 1 & 127\\ 
 & 4 & 126\\ 
 $E_{10}$ & 1 & 136\\ 
 & 4 & 137\\ 
\bottomrule
\end{tabular}
\caption{\label{tab:learning} 
%on the server
}
\end{table}

@@

\subsubsection{Abstraction}
Here, we test the benefit of different kind of tactic abstraction.
All experiments in this table are using orthogonalization.

>>>>

\todo{JU review: Still feels more like an intro than an abstract. Could be 
shorter and punchier.}
\begin{abstract}
Formalization of mathematics in interactive theorem provers (ITPs) 
relies today on many proof automation methods.
Some of these methods are general-purpose and with enough time they can prove any 
theorem. Many formalized proof scripts however also
take advantage of specialized rules and theory-based strategies. 
In the HOL4 ITP system, practically all of them are implemented as \emph{tactics}.

@@

We bridge this gap by implementing our tactical prover TacticToe on top 
of the HOL4 ITP. 
TacticToe first learns from many proof scripts
the relevance of tactics in particular proof situations.
This learned knowledge is then used in a Monte Carlo tree search algorithm that explores the most promising tactic-level proof paths. 
On a single CPU, with a time limit of 60 seconds, TacticToe proves 66.2\% 
of 7164 theorems of the standard library where as 
Eprover solves 34.5\%. The success rate rises to 69.0\% by combining the 
result of both provers.
\end{abstract}

\section{Introduction}
\tableofcontents

\todo{TG : comparison with hammers in usability}
Many of the state-of-the-art interactive theorem provers (ITPs) such as
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight}, 
  \isabelle~\cite{isabelle} 
  and \coq~\cite{coq-book}
  provide high-level parameterizable tactics for constructing the
  proofs.  Such tactics typically analyze the current goal state and
  assumptions, apply nontrivial proof transformations, which get
  expanded into possibly many basic kernel-level inferences or significant 
  parts of the proof term.
%\marginpar{not Coq}
  In this
  work we develop a tactic-level automation procedure for the \holfour ITP
  which guides selection of the tactics by learning from previous
  proofs.  Instead of relying on translation to first-order automated
  theorem provers (ATPs) as done by the hammer 
  systems~\cite{hammers4qed,tgck-cpp15}, the technique
%\marginpar{too many cites?}
%BlanchetteGKKU16,holyhammer,mizAR40
  directly searches for sequences of tactic applications that lead to
  the ITP proof, thus avoiding the translation and
  proof-reconstruction phases needed by the hammers.

\todo{TG: rewrite the plan}
  To do this, we \emph{extract and record} tactic invocations from the ITP 
  proofs (Section~\ref{sec:recording}) and
  \emph{build efficient machine learning classifiers} based on such training
  examples (Section~\ref{s:prediction}).  The learned data serves as a 
  guidance for a \emph{Monte Carlo tree search} that explores the different 
  proof paths (Section~\ref{sec:proofsearch}). The result, if
  successful, is a certified human-level proof composed of \holfour
  tactics.  The system is evaluated on a large set of theorems originating
  from \holfour (Section~\ref{s:experiments}), and we show that the performance 
  of the single best \tactictoe strategy exceeds the performance of a hammer 
  system used with a single strategy and a single efficient external prover.

\subsection{The problem}
How a user works with an ITP and how this gives rise to a machine learning problem.

@@

Input of the prediction: goal

@@

\todoi{add prediction of 4 kinds of things}
\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern 
matching constructions, opening of submodules are supported. The different 
recording steps (parsing, unfolding and replaying) are regrouped into a single 
\sml function. So, our recording is more complete and fully automatized.
\item Two types of arguments (list of theorems and terms) can be predicted 
independently of the tactic. Combinatorial explosion in the number of tactics 
is prevented through the introduction of abstracted tactics. As desired, their 
instantiations is dependent on the target goal.
\item MCTS (Monte Carlo tree search) now replaces A* in our proof search 
algorithm. MCTS gives a more balanced feedback to guide proof search by 
comparing subtrees instead of leafs. The policy and evaluation are learned 
through supervised learning.

@@

\section{Prediction}\label{s:prediction}
The learning-based selection of relevant lemmas significantly improves the 
automation for hammers~\cite{BlanchetteGKKU16}. In \tactictoe, we adapt 
one of the strongest lemma selection 
methods: the modified distance-weighted \emph{$k$ nearest-neighbour} 
classifier~\cite{ckju-pxtp13,DudaniS76} to predict three different objects: 
 goals, theorems and list of goals.

We first introduce
specificities associated with the prediction of each object: the dataset from
which objects are selected and the different use cases for the 
prediction. We then details how we predict objects suitable for the current 
goal by comparing their features with the feature of the goal.

\paragraph{Tactics}
Given a goal $g$, the classifier selects a list of input goals
ordered by their similarity to $g$. Since each input goal is associated 
with a tactic in our database, the ordered list of goals induces a ordered  
list of associated tactics which gives us a prior policy function for  MCTS. 
And tactics are tried in this order during the 
proof search. Therefore, to order the tactics correctly, it is crucial to 
estimate the similarity between goals with mathematically relevant features.

@@

Whenever we mention that we want to predict a goal, we 
actually intend to predict its associated tactics.
We derive from tactic selection a prior policy for 
our search algorithm. Lists of goals (products of tactics) evaluate the success 
of the policy and adapt it accordingly.

\paragraph{Theorems}
Because we may augment the dataset of tactics with abstracted tactics where 
list of theorems have been erased, we need to find a way to instantiate them.

This algorithm is also useful for selecting premise to ATPs. (see 
Section~\ref{sec:atp})

Given a open goal $g$, the classifier selects a list of theorems ordered by 
their 
similarity to $g$.
A dataset of theorems is constructed from the \holfour database and local 
theorem values.
We predict theorems most similar to the goal for instantiate list of theorems 
that appears as arguments of abstracted tactics (see Section 
\ref{sec:synthesis}). A list of 
predicted theorems 
are also given as input to ATPs. The number of predictions is 16 for tactic 
arguments and \metis, but is increased to 128 for \eprover.

\paragraph{List of goals}
The dataset of learned list of goals is compiled during orthogonalization.
This set is separated into positive and negative examples.
Given a list of goals outputted by a tactic,
we compute the closest lists of goals in our dataset. The ratio of positive 
examples from the total number of examples gives us a prior evaluation function 
for MCTS (see Section~{}).

\subsection{Feature Extraction}\label{sec:features}
The four objects considered are different representation of mathematical 
formulas. Therefore, we start by describing which features we extract from the 
simplest representation terms. From there, we extend the extraction mechanism 
to goals, theorems and lists of goals. Duplicated features are always removed 
so each object has an associated set of features.

@@

@@

Making prediction for an object $o$ consists sorting a set of objects 
by their similarity to $o$.

We use $sim_1$ for goals and theorems and $sim_2$ for list of goals.
The similarity $sim_1$ is used for predicting a tactic (via goals) because if a
tactic is effective on large goal $g$ then it is often suitable for a goal $g'$ 
containing sub-formulas of $g$. The same monotonicity heuristic can justify the 
use of $sim_1$ in theorems. Indeed, in practice, a large theorem is often a 
conjunction of small theorems from the same domain. And if a conjunction 
contains a formula related to a problem, then the other conjuncts 
from the same domain may also help.

\subsection{Preselection}\label{sec:dependencies}

The predictions algorithm is called many times during proof search to chose 
suitable tactics, instantiate tactic arguments by terms and theorems and
estimate the provability of a list of goals.
In order to speed up the predictions, we preselect 500 features vectors of 
tactics, theorems and list of goals.

We first rank feature vectors by their distance to the conjecture. But better 
ranked vectors only contains symbols from 
the conjecture and may not be suitable for later stages of the proof where the 
goals may have derived significantly from the conjecture. To solve this 
problem, we reorder the list so that descendants (see 
Definition~\ref{def:recdep}) of vectors inherits the similarity of their 
parent.

@@

@@

We made argument predictors for one type of arguments: list 
of theorems. We choose this very common type because perfect instantiation of 
those arguments would lead to the largest improvements (see 
Section~\ref{sec:perfect_exp}). We plan to 
extend our algorithm to term which is another frequent argument types.
But term prediction have been so far proven to weak to positively 
influence the proof search.

@@

\subsection{Search Tree}

@@

Maybe should move this definition in static description.

@@

The second case happens very rarely (less than 10 times when evaluating 7088 
theorems) and only at the beginning of the library when the learned dataset is 
small. So in our experiments, we just give the number of proof found relative 
to the number of proof attempts.
\end{definition}

\subsection{ATP Integration}~\label{sec:atp}
General-purpose proof automation mechanisms which combine proof translation to
ATPs with machine learning (``hammers'') have become quite successful in
enhancing the automation level in proof assistants~\cite{hammers4qed}.
As external automated reasoning techniques sometimes outperform the combined 
power of tactics, we would like to combine the \tactictoe search with 
stronger general purpose provers such as the one in \holyhammer for 
\holfour~\cite{tgck-cpp15}.

During our experiments,
We chose to distinguish two these general purpose tactic \metis and \eprover.
\metis is integrated in \holfour, so is already used by the default \tactictoe 
but we give it a boost as we are confident it is a strong tactic. The boos 
consits in ...

can be call with frequently with low 
timeout.
And the strongest external provers according to past experiments \cite{tgck-cpp15}.
Where as...
external calls to \eprover are computationally expensive (5 seconds at a 
minimum +  128 premises), we  
experiment with a small hammer tactic comprised of: a faster premise 
selection algorithm and
a short call (0.1 seconds + 16 premises optimized by the previous paper) to the 
internal prover \metis~\cite{metis} or a
longer call to external prover \eprover.

@@

@@

\begin{table}[ht]
\centering\ra{1.3}
\small
\begin{tabular}{ll}
\toprule
 $c_{policy}$ & Solved \\
\midrule
 $2/3$ & 120\\
 $1/2$ & 125\\
 $1/3$ & 117\\
% 1/2 & Productive & No & $E_{none}$ & 1 & 120\\
% 1/2 & Productive & No & $E_{productive}$ & 1 & 127\\
\bottomrule
\end{tabular}
\caption{\label{tab:cost_param} 
%on a local machine
}
\end{table}

@@

This experiments are performed with orthogonalization on and policy coefficient 
of $1/2$.

\begin{table}[ht]
\centering\ra{1.3}
\small
\begin{tabular}{lll}
\toprule
 PriorEvaluation & $c_{exploration}$ & Solved \\
\midrule
% 1/2 & All & No & $E_{none}$ & 1 & 122\\
% 1/2 & All & Yes & $E_{none}$ & 1 & 128\\
 none & 1 & 128\\
 productive & 1 & 134\\ 
 & 4 & 130\\
 radius = 1 & 1 & 127\\ 
 & 4 & 126\\ 
 radius = 10 & 1 & 136\\ 
 & 4 & 137\\ 
\bottomrule
\end{tabular}
\caption{\label{tab:search_parameters}}
%on the server
\end{table}

@@

\subsubsection{Orthogonalization Abstraction}
Here, we test the benefit of different kind of tactic abstraction.
All experiments in this table are using orthogonalization.
